Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands

### 2023-05-19 manuscript.Rmd
Update manuscript

The aim of this study is to identify the best combinations of seed mixtures and substrates for vital and species-rich grasslands on north- and south-exposed dike slopes. An experiment was set up to test different substrate depths, sand admixtures, seed densities, and mixture types. We expected a better development of dry grassland in the south exposition with shallow and sandy substrates, and of mesic meadows in north exposition on less sandy and deeper substrates. For steep slopes, e.g., on dikes, high seed densities are recommended for successful vegetation establishment [@KleberLerchbaumer.2017], albeit without experimental evidence.

@@

3.  Reduced soil fertility benefit target species of species-rich grasslands.

4.  High seed densities increase the likeliness of the establishment of sown plants and suppress non-target species compared to low density sowing.

@@

The target vegetation types were typical grassland types for Central Europe: lowland mesic hay meadows and semi-dry calcareous grassland [EUNIS codes: R22, R1A, @Chytry.2020; Arrhenatherion elatioris and Cirsio-Brachypodion pinnati according to the EuroVegChecklist: CM01A, DA01B, @mucina2016]. The species pool for seed mixtures of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds [Co. Krimmer, Pulling, source area 16, @prasse2010]. From these species pools, 20 species were selected for each plot in a stratified, randomised manner (Appendix A2). The aim of these random subsamples was to test types of seed mixtures and not a certain species composition. Each mixture contained seven grasses (60wt% of total seed mixture), three legumes (5%) and ten further non-legume forbs (35%) (Table \@ref(tab:seedmix)). The hay-meadow mixtures had higher community-weighted means (CWM) for specific leaf area (SLA), lower for seed mass, and higher for canopy height than the dry-grassland mixtures (Appendix A3). The south-exposed plots were sown in mid-April 2018 and the north exposition 14 days later. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions. In late-April 2018 due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks due to unsatisfactory effects on seedling emergence. The management started with a cut at 20 cm height without hay removal in August 2018, followed by standard deep cuts with hay removal in July 2019 and 2020. The surrounding area of the plots was mown thrice a year and the first time before flowering in May.

@@

@@

To compare the restoration outcomes with real references and not solely with seed mixtures, vegetation surveys were extracted from sPlotOpen [@Sabatini.2021] and our own surveys on the Danube dikes in the surroundings [@bauer2023]. We selected six dry grassland plots [EUNIS code R1A, @Chytry.2020] within SE Germany from sPlotOpen and 98 plots of our own survey, which included also hay meadows (R22), and as a negative reference ruderal, dry and anthropogenic vegetation (V38).

@@

A non-metric multidimensional scaling ordination (NMDS) with Sørensen dissimilarity was used to visualise variation in species composition in space and time. Seven species were excluded because they had an accumulated cover over all plots of\<0.5%. Finally, 343 species were included in the ordination.

@@

The ordination showed the species composition of seed mixtures and the development of the plots during four years (Figure \@ref(fig:nmds)). The NMDS confirmed that the seed mixtures were variable, albeit distinctive for hay meadows and dry grasslands and confirming the intended direction of the vegetation development. As one exception, hay meadows in south exposition did not develop towards their seed-mixture compositions.

@@

We could identify a statistically clear positive effect of the sand admixture on the persistence of sown species and on the recovery rate, but no effects of seed density or substrate depth (Figure \@ref(fig:effects)). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed positively after one year, while the recovery rate slowed down in the following years. Both expositions revealed similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong. Persistence of both seed mixture types was slightly positively affected by sand admixture in north exposition (+ 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

The vegetation developed generally in the desired direction but was still distinct from positive references and seed mixtures after four years. In the south exposition, the plots were rather similar to the negative reference of dry ruderal vegetation. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural reference grassland [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018] and five years might be too short for the assembly of secondary grasslands [cf. @nerlekar2020]

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral component need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to a study in France, sand admixture reduced nutrient loads and led to higher persistence of sown species while an increase from 25% to 50% admixture did not further increase this effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the 4^th^ year of restoration was rather small. The Favourable Conservation Status (FCS) was hardly affected by the sand admixture, which corresponds to an experiment in a quarry [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or FCS, similarly to earlier studies [@husicka2003; @baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010] or a thin substrate layer of \<15 cm, since most roots occur in the topsoil on dikes [@vannoppen2016]. Seed density had also no clear effect on persistence and FCS which fits the results of @Kaulfuss.2022, who found that a certain amount of seeds is necessary for a successful establishment of target species, but higher densities do not further improve the outcome, but rather have a slightly negative effect.

@@

Our results suggest that adapted seed mixtures can increase restoration effectiveness by sowing hay meadows in the north but not necessarily in south exposition of dikes. Furthermore, the reduction of the nutrient load through sand admixture was positive, albeit with small effect size. The question remains if sand admixture is the most efficient restoration measure to promote diversity on dikes. Increasing seed density on dike slopes does not appear to be necessary, and soil depths of 30 cm are not adverse compared to 15 cm thick substrates.

>>>>

The aim of this study is to identify the best combinations of seed mixtures and substrates for vital and species-rich grasslands on north- and south-exposed dike slopes. An experiment was set up to test different substrate depths, sand admixtures, seed densities, and mixture types. We expected a better development of dry grassland in the south exposition with shallow and sandy substrates, and of mesic meadows in north exposition on less sandy and deeper substrates. In general, we expect a better development on sandy and shallow substrates because nutrient availability is reduced [@baer2004]. For steep slopes, e.g., on dikes, high seed densities are recommended for successful vegetation establishment [@KleberLerchbaumer.2017], albeit without experimental evidence.

@@

3.  High seed densities and reduced soil fertility increase the likeliness of the establishment of sown plants and suppress non-target species compared to low density sowing.

@@

The target vegetation types were typical grassland types for Central Europe: lowland mesic hay meadows and semi-dry calcareous grassland [EUNIS codes: R22, R1A, @Chytry.2020; Arrhenatherion elatioris and Cirsio-Brachypodion pinnati according to the EuroVegChecklist: CM01A, DA01B, @mucina2016]. The species pool for seed mixtures of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds [Co. Krimmer, Pulling, source area 16, @prasse2010]. From these species pools, 20 species were selected for each plot in a stratified, randomised manner (Appendix A2). The aim of these random and unique subsamples was to test types of seed mixtures and not a certain species composition. Each mixture contained seven grasses (60wt% of total seed mixture), three legumes (5%) and ten further non-legume forbs (35%) (Table \@ref(tab:seedmix)). The hay-meadow mixtures had higher community-weighted means (CWM) for specific leaf area (SLA), lower for seed mass, and higher for canopy height than the dry-grassland mixtures (Appendix A3). The south-exposed plots were sown in mid-April 2018 and the north exposition 14 days later. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions. In late-April 2018 due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks due to unsatisfactory effects on seedling emergence. The management started with a cut at 20 cm height without hay removal in August 2018, followed by standard deep cuts with hay removal in July 2019 and 2020. The surrounding area of the plots was mown thrice a year and the first time before flowering in May.

@@

To compare the restoration outcomes with real references and not solely with seed mixtures, vegetation surveys were extracted from sPlotOpen [@Sabatini.2021] and our own surveys on the Danube dikes in the surroundings [@bauer2023]. We selected six dry grassland plots [EUNIS code R1A, @Chytry.2020] within SE Germany from sPlotOpen and 98 plots of our own survey, which included dry grasslands (*n* = 15), hay meadows (R22, *n* = 59), and as a negative reference ruderal, dry and anthropogenic vegetation (V38, *n* = 10).

@@

@@

A non-metric multidimensional scaling ordination (NMDS) with Sørensen dissimilarity (presence--absence data) was used to visualise variation in species composition in space and time. Seven species were excluded because they had an accumulated cover over all plots of \<0.5%. Finally, 343 species were included in the ordination.

@@

The ordination showed the species composition of seed mixtures and the development of the plots during four years (Figure \@ref(fig:nmds); 2D-stress: 0.21). The NMDS confirmed that the seed mixtures were variable, albeit distinctive for hay meadows and dry grasslands and confirming the intended direction of the vegetation development. As one exception, hay meadows in south exposition did not develop towards their seed-mixture compositions.

@@

A statistically clear positive effectof the sand admixture was identified on the persistence of sown species and on the recovery rate, but no effects of seed density or substrate depth (Figure \@ref(fig:effects); Persistence: *R*<sup>2</sup><sub>m</sub> = 0.86, *R*<sup>2</sup><sub>c</sub> = 0.89; Recovery: *R*<sup>2</sup><sub>m</sub> = 0.90, *R*<sup>2</sup><sub>c</sub> = 0.92; FCS: *R*<sup>2</sup><sub>m</sub> = 0.81, *R*<sup>2</sup><sub>c</sub> = 0.85). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed positively after one year, while the recovery rate slowed down in the following years. Both expositions revealed similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong. Persistence of both seed mixture types was slightly positively affected by sand admixture in north exposition (+ 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

The vegetation developed generally in the desired direction. The development after one year was very fast and afterwards very slow. For practitioners, it would be more cost-efficient to start monitoring not in the year of seeding because ruderal species are too dominant for a sensible interpretation of the results. After four years, the vegetation was still distinct from positive references and seed mixtures. In the south exposition, the plots were rather similar to the negative reference of dry ruderal vegetation typical of grassland restoration when perennial species are still developing to become dominant [@eckhoff2023]. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural reference grassland [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018] and five years might be too short for the assembly of secondary grasslands [cf. @nerlekar2020]. The annual *B. hordeaceus*, seeded in autumn 2018 as nursery plant, decreased but was still present in 179 of 288 plots in 2021.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral component need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to @chenot-lescure2022, sand admixture reduced nutrient loads and led to higher persistence of sown species while an increase from 25% to 50% admixture did not further increase this effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the 4^th^ year of restoration was rather small. The Favourable Conservation Status (FCS) was hardly affected by the sand admixture, which corresponds to an experiment in a quarry [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or FCS, similarly to earlier studies [@husicka2003; @baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010] or a thin substrate layer of \<15 cm, since most roots occur in the topsoil on dikes [@vannoppen2016]. Seed density had also no clear effect on persistence and FCS which fits the results of @Kaulfuss.2022, who found that a certain amount of seeds is necessary for a successful establishment of target species, but higher densities do not further improve the outcome, but rather have a slightly negative effect.

@@

Our results suggest that adapted seed mixtures can increase restoration effectiveness by sowing hay meadows in the north but not necessarily in south exposition of dikes. Furthermore, the reduction of the nutrient load through sand admixture was positive, albeit with small effect size. The question remains if sand admixture is the most efficient restoration measure to promote diversity on dikes. Increasing seed density on dike slopes does not appear to be necessary which contradicts common recommendations, and soil depths of 30 cm are not adverse compared to 15 cm thick substrates.
